---
title: "Module 4 — Model Checks, Comparisons and Stacking"
subtitle: "Convergence is not fit · posterior predictive checks · test statistics and p-values · KL divergence and elpd · AIC, DIC, WAIC, LOO · comparing models · stacking · the estimand from draws"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-compute");</script>
```

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
set.seed(540)
```

After Module 3 you can fit a model and show that the MCMC converged. That only means you have an accurate picture of the posterior **under the model you assumed**. A badly wrong model converges just as happily as a good one. Module 4 asks the next questions: *Is this model any good? Where does it fall short? Would another model do better?*

The first tool is the **posterior predictive check (PPC)**. Because Bayesian models are generative, you can simulate "replicated" datasets from the fitted model and compare them with the data you actually observed. If the replications systematically miss a feature you care about (the tails, the share of zeros, the pattern across subgroups), the model needs work.

The second tool estimates **out-of-sample predictive performance**. The ideal measure is the expected log predictive density for new data (elpd), rooted in the Kullback–Leibler divergence. AIC, DIC, WAIC and leave-one-out cross-validation (LOO) are increasingly good estimates of it. brms reports LOO with a standard error, so you can see whether a difference between models is larger than its uncertainty.

The third idea is **stacking**: rather than picking one model, average the predictive distributions of several, with weights chosen for predictive performance. The module ends by computing the actual target of inference, often a non-linear function of parameters, from the draws.

::: {.callout-note collapse="true"}
## Sources, reading and seminar
**Course notes:** Module 4 "Model checks, comparisons and stacking" (pp. 163–216). Reading deadline: **Seminar 4, Monday 28 September 2026** (topic: PPC diagnostics, stacking).

**Optional:** McElreath, *Statistical Rethinking*, chapter 7 (overfitting, information criteria, LOO). Gabry et al. (2019), "Visualization in Bayesian workflow", *JRSS A*, the source for `pp_check()` graphics. Vehtari, Gelman and Gabry (2017), "Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC", *Statistics and Computing*.

**Examples on this page** (length of hospital stay, sodium readings with faulty values, flu vaccination by age) are written for this app. The course notes' own example (school meals and numeracy) is in the notes.
:::

::: {.callout-tip collapse="true"}
## Before you start: three things you need
**Posterior predictive distribution** ([1.6](module1.qmd#t1-6)): draw θ from the posterior, then simulate data from $p(y \mid \theta)$. One such simulated dataset of the same size as your data is a *replication* $y^\text{rep}$.

**Log-likelihood and deviance.** Deviance = −2 × log-likelihood. Higher log-likelihood (lower deviance) means the model gives the data more probability. Adding parameters never lowers the in-sample fit of nested models, which is why complexity must be penalised.

**Overfitting.** A model tuned too closely to one sample fits that sample well and new data badly. In-sample fit is optimistic; out-of-sample fit is what matters for prediction.
:::

# Part A · Concepts

## 4.1 Convergence is not fit {#t4-1 .topic flag="ilo"}

Two separate questions:

| Question | Tool | Module |
|---|---|---|
| Does my MCMC sample represent the posterior *of this model*? | Trace plots, R-hat, ESS | 3 |
| Is *this model* a reasonable description of the data, for my purpose? | Posterior predictive checks, LOO | 4 |

Fit a normal model to data from a heavy-tailed distribution and the chains converge perfectly: R-hat 1.00, thousands of effective draws. Convergence is a property of the algorithm; fit is a property of the model. Check convergence first (otherwise the checks of fit use a wrong posterior), then check fit **before interpreting estimates**.

::: {.callout-warning}
## Common confusion: "no warnings from brms, so the model is fine"
brms warns about sampling problems (divergences, R-hat, ESS). It never warns that the likelihood is wrong, that an important predictor is missing, or that the prior is absurd. Those need posterior (and prior) predictive checks and subject-matter thinking.
:::

## 4.2 Graphical posterior predictive checks {#t4-2 .topic flag="ilo project"}

1. For many posterior draws $\theta^{(s)}$, simulate a replicated dataset $y^{\text{rep}(s)}$ of the same size and with the same covariates as the observed data.
2. Compare the replications with the observed $y$, graphically or through a statistic.

In brms: `pp_check(fit)` overlays the density of the observed outcome (dark line) on densities of, by default, 10 replications (light lines). `ndraws = 50` shows more.

```{r}
#| label: fig-ppc
#| fig-cap: "Posterior predictive checks for simulated lengths of hospital stay (days). Left: a normal model: replications are symmetric and go below zero. Right: a normal model for log(stay), back-transformed: replications resemble the data."
#| fig-height: 3.2
stay <- round(rlnorm(80, meanlog = log(4), sdlog = 0.7), 1)
n <- length(stay)
draw_normal_posterior <- function(y, S) {
  sigma2 <- (length(y) - 1) * var(y) / rchisq(S, df = length(y) - 1)
  data.frame(mu = rnorm(S, mean(y), sqrt(sigma2 / length(y))), sigma = sqrt(sigma2))
}
post_raw <- draw_normal_posterior(stay, 20)
post_log <- draw_normal_posterior(log(stay), 20)
reps <- rbind(
  do.call(rbind, lapply(1:20, function(s) data.frame(
    y = rnorm(n, post_raw$mu[s], post_raw$sigma[s]), rep = s, model = "Normal model for stay"))),
  do.call(rbind, lapply(1:20, function(s) data.frame(
    y = exp(rnorm(n, post_log$mu[s], post_log$sigma[s])), rep = s, model = "Log-normal model for stay")))
)
obs <- rbind(data.frame(y = stay, model = "Normal model for stay"),
             data.frame(y = stay, model = "Log-normal model for stay"))
reps$model <- factor(reps$model, levels = c("Normal model for stay", "Log-normal model for stay"))
obs$model <- factor(obs$model, levels = levels(reps$model))
ggplot() +
  geom_density(data = reps, aes(y, group = rep), colour = "#9ec5e8", linewidth = 0.3) +
  geom_density(data = obs, aes(y), colour = "#0b3c6b", linewidth = 1.1) +
  facet_wrap(~ model) + coord_cartesian(xlim = c(-8, 25)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(x = "length of stay (days)", y = "density")
```

Each replication differs from the others, so no replication will match the data exactly. The question is whether the observed data look like **one more replication**. On the left, they do not: the observed distribution is skewed, has no negative values and peaks earlier than any replication.

**Other useful graphical checks** (all `pp_check(fit, type = ...)`):

| `type =` | Shows | Good for |
|---|---|---|
| `"dens_overlay"` (default) | densities of $y$ and replications | continuous outcomes |
| `"hist"`, `"ecdf_overlay"` | histograms, cumulative distributions | tails, discreteness |
| `"stat"`, `"stat_2d"` | a statistic of $y$ vs its replication distribution | any outcome, incl. binary |
| `"stat_grouped"` | the statistic within each level of a grouping variable | subgroups, clusters (the variable must be in the model; otherwise use `bayesplot::ppc_stat_grouped()`, Task 4.2) |
| `"intervals"`, `"scatter_avg"` | predictive intervals vs observed per observation | outliers, regression fit |
| `"bars"` | counts per category | ordinal and count outcomes |

## 4.3 Test statistics and posterior predictive p-values {#t4-3 .topic flag="ilo project"}

For a binary outcome, a density plot of zeros and ones says little. Choose a **test statistic** $T(y)$ instead: the proportion of ones, the mean in a subgroup, the maximum, the share of zeros, a correlation between repeated measurements. Compute it for the data and for every replication:

$$p_\text{B} = P\big(T(y^\text{rep}) \geq T(y) \mid y\big) \approx \frac{1}{S}\sum_s I\big(T(y^{\text{rep}(s)}) \geq T(y)\big).$$

- A value near **0.5** means the observed statistic sits in the middle of the replications: the model reproduces this feature.
- Near **0 or 1** means the observed value is in a tail: the model misses this feature. The notes suggest treating values below about 0.11 or above 0.89 as a reason to extend the model (the threshold depends on how important the feature is).
- **Choose statistics that the model does not fit directly.** A model with an intercept will always reproduce the overall mean, so checking the mean is weak. Check what the model does *not* explicitly fit: tails, subgroups, variables left out of the model, the dependence between repeated measurements.

:::: {.quiz}
A logistic regression includes an intercept and age. Which posterior predictive check is most likely to reveal a problem?

- The overall proportion of ones
- **The proportion of ones within each region**
- The mean of the replicated proportion of ones
- The number of observations in each replication

::: {.exp}
The intercept makes the model reproduce the overall proportion almost by construction. A grouping variable that is not in the model (region) is where systematic misfit can show up. The last two options are not checks of fit at all.
:::
::::

::: {.callout-warning}
## Common confusion: PPCs and multiple testing
Posterior predictive checks are not inferences about parameters or predictions of new data: they are diagnostics, like trace plots. The notes say there is no need to correct for multiple testing. Check as many features as are relevant, then act: extend the model, or report the limitation.
:::

## 4.4 Choosing checks for your question {#t4-4 .topic flag="project"}

Which features matter depends on the research question. Some starting points:

| Data structure | Useful checks |
|---|---|
| Continuous outcome | density overlay; skewness; min and max; share of extreme values |
| Binary outcome | proportion of ones overall and **by subgroup** (`stat_grouped`), including groups *not* in the model |
| Counts | share of zeros; variance vs mean (overdispersion); max |
| Clustered or repeated measures | cluster means (`stat_grouped` by cluster); SD of cluster means; correlation between time points; scatter of $y_{i2}$ against $y_{i1}$, observed vs replicated |
| Items in a questionnaire | item-specific means; log odds ratios between pairs of items |

A failed check points to a specific extension: heavier tails (Student-t), a transformation, a quadratic term or spline, an interaction, a random effect, an overdispersion parameter.

## 4.5 Robust likelihoods: normal or Student-t? {#t4-5 .topic}

When a few values are wildly off (transcription errors, faulty samples, a minority of unusual patients), a normal likelihood lets them inflate $\sigma$ and pull the mean. A **Student-t** likelihood has heavier tails, so outliers are less "surprising" and have less leverage. In brms: `family = student()`, with the degrees of freedom $\nu$ estimated (default prior `gamma(2, 0.1)`) or fixed via `bf(y ~ 1, nu = 4)`.

- Small $\nu$ (3–5): heavy tails. Large $\nu$ (> 30): practically normal.
- The Student-t is not the true model for contaminated data (a mixture of good and bad readings), but it is usually **less wrong** than the normal. Compare them with PPCs and LOO (Task 4.1).
- $\sigma$ in a Student-t model is a *scale*, not the SD: $\text{SD} = \sigma\sqrt{\nu/(\nu-2)}$ for $\nu > 2$.

## 4.6 Out-of-sample performance: KL divergence and elpd {#t4-6 .topic}

The objective way to judge a model is how well it predicts **new** data. The notes build the ideal measure from the **Kullback–Leibler divergence** between the true distribution of new data $g$ and the model's predictive distribution $f$:

$$D_\text{KL}(g \,\|\, f) = \int g(\tilde y) \log\frac{g(\tilde y)}{f(\tilde y)}\,d\tilde y \;\geq 0, \qquad = 0 \text{ only if } f = g.$$

- It is **asymmetric**: $D_\text{KL}(g\|f) \neq D_\text{KL}(f\|g)$. A too-narrow prediction that misses where data actually land is punished more than a too-wide one ("the expected surprise" when reality turns out to be $g$).
- When comparing models, $\int g \log g$ is the same constant for all of them, so only $E_g[\log f(\tilde y)]$ matters. Summed over observations this is the **expected log pointwise predictive density**:

$$\text{elpd} = \sum_{i=1}^n E_{g_i}\big[\log p(\tilde y_i \mid y)\big].$$

Higher elpd = better predictions. We never know $g$, so every method below is an *estimate* of elpd.

## 4.7 AIC, DIC and WAIC {#t4-7 .topic}

All three take the in-sample fit and subtract a **penalty for complexity** to correct its optimism. On the deviance scale: criterion = −2 × (estimated elpd); **lower is better**.

| Criterion | In-sample fit uses… | Penalty | Notes |
|---|---|---|---|
| AIC | log-likelihood at the MAP/MLE (one point) | $k$ = number of parameters | ignores posterior uncertainty; assumes a roughly normal posterior |
| DIC | average deviance over posterior draws | posterior variance of the deviance / 2 (one version) | uses the posterior, still assumes near-normality |
| **WAIC** | $\text{lppd} = \sum_i \log\Big(\frac1S\sum_s p(y_i \mid \theta^{(s)})\Big)$ | $p_\text{WAIC} = \sum_i \text{var}_s\big(\log p(y_i \mid \theta^{(s)})\big)$ | pointwise, fully Bayesian, no normality assumption |

$\text{elpd}_\text{WAIC} = \text{lppd} - p_\text{WAIC}$, and $\text{WAIC} = -2\,\text{elpd}_\text{WAIC}$. The penalty $p_\text{WAIC}$ is an "effective number of parameters": with strong priors or hierarchical shrinkage it is smaller than the raw parameter count.

Because WAIC (and LOO) are **sums over observations**, you get a standard error from the pointwise terms, and a paired standard error for the difference between two models (Section 4.9).

## 4.8 Leave-one-out cross-validation {#t4-8 .topic flag="project"}

Cross-validation estimates elpd directly: fit on part of the data, evaluate the log predictive density on the held-out part. No penalty is needed, because the test data were not used for fitting.

- **k-fold CV**: split into $k$ folds, fit $k$ times. Each fit uses a fraction $(k-1)/k$ of the data.
- **LOO-CV**: $k = n$. Refitting $n$ times is expensive, but **PSIS-LOO** (Pareto-smoothed importance sampling) approximates it from a single fit: `loo(fit)` in brms.
- **Pareto $\hat k$ diagnostics.** Each observation gets a $\hat k$. Values below 0.7 are fine. Above 0.7, that observation is highly influential and the approximation is unreliable for it: use `loo(fit, moment_match = TRUE)` (the model must be fitted with `save_pars = save_pars(all = TRUE)`), `loo(fit, reloo = TRUE)`, or `kfold(fit, K = 10)`.

**What does "leave one out" mean for your data?** LOO answers "how well do I predict a new observation *from a group I have already seen*?" To predict a **new group** (a new hospital, region or country), leave out whole groups: `kfold(fit, folds = "grouped", group = "hospital", K = 10)`. For time series, predict the future from the past. The notes recommend LOO-PSIS or k-fold over WAIC; LOO also gives the $\hat k$ warnings that WAIC lacks.

## 4.9 Comparing models {#t4-9 .topic flag="project"}

```r
loo_compare(loo(fit_a), loo(fit_b))
#>        elpd_diff se_diff
#> fit_b    0.0       0.0
#> fit_a  -12.3       4.1
```

- The **best model is on the top row**, with `elpd_diff = 0`. Other rows show how much worse they are, with the standard error of the *paired* difference.
- **Rules of thumb** (notes, based on Stan's LOO FAQ): trust a difference only if $|\text{elpd\_diff}| > 4$ **and** it is larger than about 2 × `se_diff`. Above: difference −12.3 with SE 4.1 → about 3 SE, so a real advantage for `fit_b`.
- Works for **non-nested models** (logit vs probit, normal vs Student-t) and different likelihoods, as long as they predict **the same outcome on the same scale**. A model for log(y) must include the Jacobian to be compared with a model for $y$ (brms handles this when you model `lognormal()` rather than `log(y)`).
- The comparison includes the **prior**: a model is "model + prior".

:::: {.quiz}
`loo_compare` shows elpd_diff = −2.1 and se_diff = 1.8 for model A vs model B (B on top). What do you conclude?

- B predicts clearly better, so choose B
- **No clear difference: consider both, or the simpler one**
- A predicts better, because its elpd_diff is negative
- The standard error is too large, so LOO has failed

::: {.exp}
|−2.1| < 4 and about 1.2 SE: within the noise of the estimate. Report both, check whether they lead to the same conclusions, or stack them. A negative elpd_diff means worse than the top model.
:::
::::

::: {.callout-warning}
## Common confusion: model comparison as variable selection
Choosing variables by repeated model comparison is a data-driven search with the same multiple-testing problems as stepwise selection, and the chosen model's intervals are too narrow. Use comparison to learn which features predict and which patterns a model misses; choose adjustment variables from subject-matter knowledge (confounders) instead.
:::

## 4.10 Stacking and Bayesian model averaging {#t4-10 .topic}

If several models predict reasonably well, a **weighted average of their predictive distributions** often predicts better than any single one, without fitting a new model:

$$p(\tilde y \mid y, \text{stack}) = \sum_j w_j\, p(\tilde y \mid y, \text{model}_j), \qquad w_j \geq 0,\; \sum_j w_j = 1.$$

- **WAIC or LOO weights** ("pseudo-BMA", as in the notes): $w_j = \dfrac{\exp(-\tfrac12 \text{WAIC}_j)}{\sum_k \exp(-\tfrac12 \text{WAIC}_k)}$. A model a few WAIC units worse gets a small weight; one much better gets nearly all.
- **Stacking weights**: choose $w$ to maximise the LOO predictive density of the mixture. Models that are good in *different regions* of the data both get weight.
- In brms: `loo_model_weights(list(a = loo_a, b = loo_b), method = "stacking")` or `"pseudobma"`; `pp_average(fit_a, fit_b)` gives the averaged predictions.
- Sampling from the stack: take $w_j \times S$ draws from model $j$'s posterior predictive, then combine.

**Why it helps.** Two models that err in opposite directions in some region cancel part of each other's error. If one model is better everywhere, it gets (nearly) all the weight and stacking changes nothing.

**Caution.** Stacking is for **prediction**. Averaging a *parameter* across models is often meaningless, because "the coefficient of x" means something different when the other covariates differ.

## 4.11 The estimand, computed from the draws {#t4-11 .topic flag="ilo project"}

Once a model checks out, compute **the quantity that answers your research question**, which is rarely a raw coefficient:

- a **risk difference or risk ratio** at chosen covariate values (from `posterior_epred`), rather than a log odds ratio;
- a **ratio of two effects** ("the intervention is worth 0.6 years of learning" in the notes' school meals example is `b_meals / b_wave`);
- a **probability of a clinically relevant effect**, $P(\text{effect} > \delta \mid y)$;
- **ranks** or the probability of being best ([Module 5](module5.qmd#t5-11)).

Recipe: draws → compute the estimand in every draw → summarise (median, 95% interval, probabilities). Non-linear functions need no delta method: uncertainty propagates automatically.

## When to use which check {#when-to-use}

- **Always**: convergence diagnostics, then at least one graphical PPC and one or two test statistics targeted at your question.
- **Choosing between plausible models** (likelihoods, links, functional forms, random-effect structures): LOO with `loo_compare`, plus PPCs showing *where* they differ.
- **Pure prediction with several decent models**: stacking.

## Limitations {#limitations}

- PPCs use the data twice (to fit and to check), so they are conservative: they find gross misfit more easily than subtle misfit. p-values near 0.5 do not prove the model right.
- LOO estimates prediction for data *like yours*. It cannot detect confounding, and a model with better predictions is not necessarily the right causal model.
- LOO is unreliable with many high Pareto $\hat k$ values (small clusters, very flexible models); use k-fold CV then.
- Information criteria compare models on the same outcome. They cannot compare models fitted to different datasets or different transformations without adjustment.

# Part B · Practice

## In the browser: a posterior predictive check and WAIC by hand {#browser}

::: {.callout-note}
## Interactive R
Runs in your browser. A normal model with unknown mean and SD and the reference prior has an exact posterior ([2.11](module2.qmd#t2-11)), so no MCMC is needed.
:::

```{webr-r}
#| label: ppc-by-hand
set.seed(4)
length_of_stay <- round(rlnorm(80, meanlog = log(4), sdlog = 0.7), 1)   # simulated, days
n <- length(length_of_stay); S <- 4000

draw_normal_posterior <- function(y, S) {        # exact posterior, prior 1/sigma^2
  sigma2 <- (length(y) - 1) * var(y) / rchisq(S, df = length(y) - 1)
  data.frame(mu = rnorm(S, mean(y), sqrt(sigma2 / length(y))), sigma = sqrt(sigma2))
}
post_raw <- draw_normal_posterior(length_of_stay, S)        # model A: normal for stay
post_log <- draw_normal_posterior(log(length_of_stay), S)   # model B: normal for log(stay)

# One replicated dataset per posterior draw (rows = draws)
yrep_A <- t(sapply(1:S, function(s) rnorm(n, post_raw$mu[s], post_raw$sigma[s])))
yrep_B <- t(sapply(1:S, function(s) exp(rnorm(n, post_log$mu[s], post_log$sigma[s]))))

skewness <- function(y) mean((y - mean(y))^3) / sd(y)^3
ppp <- function(yrep, stat) mean(apply(yrep, 1, stat) >= stat(length_of_stay))
rbind(minimum  = c(A = ppp(yrep_A, min),      B = ppp(yrep_B, min)),
      maximum  = c(A = ppp(yrep_A, max),      B = ppp(yrep_B, max)),
      skewness = c(A = ppp(yrep_A, skewness), B = ppp(yrep_B, skewness)))

hist(apply(yrep_A, 1, min), breaks = 40, main = "Model A: replicated minimum",
     xlab = "min(y_rep)"); abline(v = min(length_of_stay), col = "red", lwd = 2)
```

1. Which statistics does model A fail? What share of model A's replicated stays are negative (`mean(yrep_A < 0)`)?
2. Invent a statistic that model B might fail, for example the share of stays longer than 14 days. Check it.

::: {.callout-tip collapse="true"}
## Model answer
1. Model A fails the minimum (p ≈ 0.001: replicated minima are almost always below the observed 1.1 days, usually negative) and skewness (p ≈ 0: the data are right-skewed, the normal is symmetric). The maximum is borderline (p ≈ 0.07). About 7% of replicated stays are negative, which is impossible. Model B's p-values are all between 0.3 and 0.8.
2. `ppp(yrep_B, function(y) mean(y > 14))` is around 0.5. Model B passes, as it should, since these data were simulated from a log-normal. With real data, look for features that matter clinically: the share of very long stays drives bed planning.
:::

```{webr-r}
#| label: waic-by-hand
# Pointwise log-likelihood: rows = posterior draws, columns = observations
loglik_A <- sapply(length_of_stay, function(y) dnorm(y, post_raw$mu, post_raw$sigma, log = TRUE))
loglik_B <- sapply(length_of_stay, function(y) dlnorm(y, post_log$mu, post_log$sigma, log = TRUE))
# dlnorm, not dnorm on log(y): both models must give densities for stay in days

waic_by_hand <- function(loglik) {
  lppd_i   <- log(colMeans(exp(loglik)))    # log of the average likelihood, per observation
  p_waic_i <- apply(loglik, 2, var)         # penalty, per observation
  elpd_i   <- lppd_i - p_waic_i
  c(elpd_waic = sum(elpd_i), p_waic = sum(p_waic_i), waic = -2 * sum(elpd_i))
}
w <- rbind(A_normal = waic_by_hand(loglik_A), B_lognormal = waic_by_hand(loglik_B))
w

# Paired difference and its standard error
elpd_i_A <- log(colMeans(exp(loglik_A))) - apply(loglik_A, 2, var)
elpd_i_B <- log(colMeans(exp(loglik_B))) - apply(loglik_B, 2, var)
difference <- elpd_i_B - elpd_i_A
c(elpd_diff = sum(difference), se_diff = sqrt(n * var(difference)))

# Pseudo-BMA weights from WAIC
weights <- exp(-0.5 * w[, "waic"]); weights / sum(weights)
```

3. Interpret the difference using the rules of thumb from Section 4.9.
4. Why are the weights so extreme?
5. Why would comparing `dnorm(log(y), ...)` for model B with `dnorm(y, ...)` for model A be wrong?

::: {.callout-tip collapse="true"}
## Model answer
3. Model B's elpd is about 20 higher, with an SE of the difference of about 5: more than 4 and about 4 SE. A clear predictive advantage for the log-normal.
4. A WAIC difference of 40 means a weight ratio of $e^{20}$: essentially all weight goes to B. Pseudo-BMA weights are extreme whenever one model is clearly better. Stacking behaves similarly here, since B is better everywhere.
5. They would be densities for different variables (days vs log-days). A density for log(y) must be divided by y (the Jacobian) to become a density for y; `dlnorm` does exactly that. Otherwise the comparison mixes units.
:::

## In RStudio: brms tasks {#rstudio}

::: {.callout-important}
## How the RStudio tasks work
Each task simulates its own data. Try first, then open the solution. The solutions were run with brms 2.23; your numbers will differ slightly. Download all Module 4 tasks: [module4-tasks.R](tasks/module4-tasks.R).
:::

::: {.rstudio-task}
Task 4.1 · Normal or Student-t for sodium readings with some faulty values

```r
set.seed(520)
bad_reading <- rbinom(300, size = 1, prob = 0.05)
sodium <- data.frame(value = ifelse(bad_reading == 1,
                                    rnorm(300, mean = 140, sd = 15),
                                    rnorm(300, mean = 140, sd = 3)))
```

1. Fit `value ~ 1` with `family = gaussian()` and with `family = student()` (priors: Intercept `normal(140, 10)`, sigma `exponential(0.2)`, nu `gamma(2, 0.1)`). Do both converge?
2. Compare `pp_check(fit, ndraws = 50)` for the two models.
3. Define `share_near <- function(y) mean(abs(y - 140) < 3)` and `share_extreme <- function(y) mean(abs(y - 140) > 15)`. Use `pp_check(fit, type = "stat", stat = "share_near")` and compute the posterior predictive p-values by hand.
4. Compare the models with `loo_compare()`.

::: {.callout-tip collapse="true"}
## Solution
```r
fit_normal <- brm(value ~ 1, family = gaussian(),
                  prior = c(prior(normal(140, 10), class = "Intercept"),
                            prior(exponential(0.2), class = "sigma")),
                  data = sodium, chains = 4, iter = 2000, seed = 520, refresh = 0)
fit_student <- brm(value ~ 1, family = student(),
                   prior = c(prior(normal(140, 10), class = "Intercept"),
                             prior(exponential(0.2), class = "sigma"),
                             prior(gamma(2, 0.1), class = "nu")),
                   data = sodium, chains = 4, iter = 2000, seed = 520, refresh = 0)
pp_check(fit_normal, ndraws = 50) + xlim(110, 170)
pp_check(fit_student, ndraws = 50) + xlim(110, 170)

yrep_normal  <- posterior_predict(fit_normal)
yrep_student <- posterior_predict(fit_student)
mean(apply(yrep_normal, 1, share_near) >= share_near(sodium$value))
mean(apply(yrep_student, 1, share_near) >= share_near(sodium$value))
# ... and the same for share_extreme
loo_compare(loo(fit_normal), loo(fit_student))
```

What we got: both converge (Rhat 1.00, ESS > 2,000). **Normal**: mean 140.08, `sigma` 4.79, inflated by the faulty readings (the good readings have SD 3). **Student-t**: mean 139.79, `sigma` 2.96, `nu` 3.7 (heavy tails).

Test statistics. Observed: 59.7% of readings lie within 3 mmol/L of 140, and 1.7% lie more than 15 away.

| | `share_near` | `share_extreme` |
|---|---|---|
| Normal model, p-value | 0.0003 | 0.0003 |
| Student-t model, p-value | 0.77 | 0.23 |

The normal model fails **both** checks: with σ = 4.8 it expects about 47% of readings near 140 (too few) and about 0.2% beyond 15 (too few). It cannot have a sharp peak *and* heavy tails at the same time. The Student-t passes both. The density overlay shows the same thing: the normal replications are too flat in the middle.

`loo_compare`: the Student-t model is on top; the normal model has elpd_diff = −55.8 with se_diff = 19.2, about 3 SE. The heavy-tailed model predicts clearly better. All Pareto $\hat k$ < 0.7.
:::
:::

::: {.rstudio-task}
Task 4.2 · A binary outcome: flu vaccination by age

```r
set.seed(521)
vaccine <- data.frame(age = round(runif(800, min = 20, max = 85)))
true_logodds <- -1.2 + 0.0008 * (vaccine$age - 20)^2       # uptake accelerates with age
vaccine$vaccinated <- rbinom(800, size = 1, prob = plogis(true_logodds))
vaccine$age10 <- (vaccine$age - 50) / 10
vaccine$age_group <- cut(vaccine$age, breaks = c(19, 35, 50, 65, 75, 85))
```

1. Fit `vaccinated ~ age10` (`family = bernoulli()`, priors `normal(0, 1.5)` Intercept and `normal(0, 1)` b). Check `pp_check(fit, type = "stat", stat = "mean")`. Is that a useful check?
2. Check the share vaccinated within each age group. Where does the model fail? Note: `pp_check(fit, type = "stat_grouped", group = "age_group")` stops with *"Variable 'age_group' could not be found in the data"*: brms only accepts grouping variables that are in the model, even with `newdata`. Call bayesplot directly instead: `bayesplot::ppc_stat_grouped(y = vaccine$vaccinated, yrep = posterior_predict(fit), group = vaccine$age_group, stat = "mean")`.
3. Fit `vaccinated ~ age10 + I(age10^2)` and repeat. Compare the two models with LOO.

::: {.callout-tip collapse="true"}
## Solution
```r
fit_linear <- brm(vaccinated ~ age10, family = bernoulli(),
                  prior = c(prior(normal(0, 1.5), class = "Intercept"),
                            prior(normal(0, 1), class = "b")),
                  data = vaccine, chains = 4, iter = 2000, seed = 521, refresh = 0)
fit_quadratic <- brm(vaccinated ~ age10 + I(age10^2), family = bernoulli(),
                     prior = c(prior(normal(0, 1.5), class = "Intercept"),
                               prior(normal(0, 1), class = "b")),
                     data = vaccine, chains = 4, iter = 2000, seed = 521, refresh = 0)
pp_check(fit_linear, type = "stat", stat = "mean")

yrep_linear    <- posterior_predict(fit_linear)
yrep_quadratic <- posterior_predict(fit_quadratic)
bayesplot::ppc_stat_grouped(y = vaccine$vaccinated, yrep = yrep_linear,
                            group = vaccine$age_group, stat = "mean")
bayesplot::ppc_stat_grouped(y = vaccine$vaccinated, yrep = yrep_quadratic,
                            group = vaccine$age_group, stat = "mean")
loo_compare(loo(fit_linear), loo(fit_quadratic))
```

What we got: the overall mean check is useless here. Both models reproduce the overall share almost by construction, and brms itself warns that "mean" is a weak statistic.

Share vaccinated per age group, with the posterior predictive p-value $P(\text{replicated share} \geq \text{observed})$:

| Age group | Observed | Linear: replicated 5%–95% | p (linear) | p (quadratic) |
|---|---|---|---|---|
| 20–35 | 0.28 | 0.15–0.28 | 0.05 | 0.34 |
| 36–50 | 0.30 | 0.31–0.45 | 0.97 | 0.79 |
| 51–65 | 0.49 | 0.49–0.63 | 0.97 | 0.73 |
| 66–75 | 0.78 | 0.62–0.78 | 0.07 | 0.09 |
| 76–85 | 0.83 | 0.72–0.87 | 0.22 | 0.73 |

The linear model shows a **systematic pattern**: too low at the youngest and at 66–75, too high in the middle. That is the signature of a curved relationship forced into a straight line on the logit scale. The quadratic model removes the pattern (one group is borderline at 0.09, which is what you expect by chance among five checks).

`loo_compare`: quadratic on top, linear −4.2 with SE 3.3. Just above 4 but only 1.3 SE: **by the rules of thumb, not a clear difference**. The PPC detected a systematic misfit that the LOO average hardly registers, because LOO sums over all 800 people and the misfit is modest for each. PPCs target the features *you* care about; LOO answers a more general question.
:::
:::

::: {.rstudio-task}
Task 4.3 · Stacking weights

Using the two LOO objects from Task 4.2, compute `loo_model_weights(..., method = "stacking")` and `method = "pseudobma"`. Do the two methods agree? When would they differ?

::: {.callout-tip collapse="true"}
## Solution
```r
loo_linear    <- loo(fit_linear)
loo_quadratic <- loo(fit_quadratic)
loo_model_weights(list(linear = loo_linear, quadratic = loo_quadratic), method = "stacking")
loo_model_weights(list(linear = loo_linear, quadratic = loo_quadratic), method = "pseudobma")
```

What we got: stacking gives the linear model 0.11 and the quadratic 0.89; pseudo-BMA+ gives 0.13 and 0.87. They agree because the quadratic model is better almost everywhere. They would differ if each model predicted better in a *different* part of the data (say, one better for the young and the other for the old): stacking would then keep both with substantial weight, while pseudo-BMA would still favour the one with the higher total elpd.
:::
:::

::: {.rstudio-task}
Task 4.4 · The estimand: how much more likely is a 70-year-old to be vaccinated than a 40-year-old?

Using the quadratic model, compute the posterior of the risk difference $P(\text{vaccinated} \mid 70) - P(\text{vaccinated} \mid 40)$ with `posterior_epred()`. Report the median, a 95% interval and $P(\text{difference} > 0.3)$.

::: {.callout-tip collapse="true"}
## Solution
```r
ages_of_interest <- data.frame(age10 = c((40 - 50) / 10, (70 - 50) / 10))
risk_draws <- posterior_epred(fit_quadratic, newdata = ages_of_interest)   # draws x 2
risk_difference <- risk_draws[, 2] - risk_draws[, 1]
quantile(risk_difference, c(0.025, 0.5, 0.975))
mean(risk_difference > 0.3)
```

What we got: median risk difference 0.40, 95% credible interval 0.33 to 0.46, and $P(\text{difference} > 0.3 \mid y) \approx 0.999$. A 70-year-old is about 40 percentage points more likely to be vaccinated than a 40-year-old. That sentence answers the question directly, which a log-odds coefficient for `age10` and another for its square do not.
:::
:::

## Why-questions {#why-questions}

::: {.callout-note collapse="true"}
## Why check features the model does not fit directly?
A model reproduces what it was built to reproduce: an intercept guarantees the overall mean, a group term guarantees the group means. Misfit shows up in features that were not built in: tails, subgroups, dependence, variables left out.
:::

::: {.callout-note collapse="true"}
## Why does WAIC need a penalty but cross-validation does not?
WAIC evaluates the model on the same data used to fit it, which is optimistic; the penalty corrects for that optimism. Cross-validation evaluates on held-out data, so there is no optimism to correct (only a small pessimism, because each fit uses fewer observations).
:::

::: {.callout-note collapse="true"}
## Why is the standard error of an elpd difference smaller than the two separate standard errors suggest?
Both models are evaluated on the same observations, so their pointwise elpd values are strongly correlated. The paired SE uses the variance of the pointwise *differences*, which removes the shared variation between observations.
:::

## Spaced review: Modules 1–3 {#review}

::: {.callout-note collapse="true"}
## 1. What is the difference between posterior_epred and posterior_predict?
epred: draws of the expected outcome (the mean, a probability). predict: draws of a new observation, which adds the outcome's own randomness. ([3.12](module3.qmd#t3-12))
:::

::: {.callout-note collapse="true"}
## 2. Your summary shows Rhat 1.00 but 50 divergent transitions. Can you trust it?
Not yet. Divergences mean the sampler may have missed a region of the posterior, so estimates may be biased. Increase adapt_delta, and consider reparameterising or tighter priors. ([3.7](module3.qmd#t3-7))
:::

::: {.callout-note collapse="true"}
## 3. Why does a Beta(1, 1) prior give credible intervals with roughly nominal coverage?
It adds no information, so the posterior is driven by the likelihood like a classical interval. Priors centred away from the truth under-cover. ([2.8](module2.qmd#t2-8))
:::
