---
title: "Module 1 — Bayesian Concepts"
subtitle: "Two meanings of probability · Bayes' rule · a whole posterior from a discrete prior · summarising a posterior · posterior predictive distributions · confidence vs credible intervals"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-found");</script>
```

```{r}
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
set.seed(510)

# Capture-recapture example used throughout the page
marked     <- 30     # fish marked and released
caught     <- 25     # fish caught a week later
recaptured <- 5      # of those, carrying a mark
pond_size  <- 50:3000                     # every possible number of fish
prior      <- dlnorm(pond_size, meanlog = log(300), sdlog = 0.6)
prior      <- prior / sum(prior)
likelihood <- dhyper(recaptured, marked, pond_size - marked, caught)
posterior  <- prior * likelihood / sum(prior * likelihood)
post_cdf   <- cumsum(posterior)
post_mean  <- round(sum(pond_size * posterior))
post_median <- pond_size[which(post_cdf >= 0.5)[1]]
post_map   <- pond_size[which.max(posterior)]
ci_lower   <- pond_size[which(post_cdf >= 0.05)[1]]
ci_upper   <- pond_size[which(post_cdf >= 0.95)[1]]
```

Every Bayesian analysis in this course is the same machine: **prior × likelihood → posterior**, and then everything you want to know (a best guess, an interval, a probability, a prediction, a decision) is read off the posterior. Module 1 shows that machine in the simplest possible setting, where the parameter can only take a list of values and Bayes' rule is just multiplication followed by division by a sum. No integrals and no MCMC yet.

The module opens with the philosophical difference that drives everything else: what *is* a probability? The classical (frequentist) definition only allows probabilities for repeatable events, so a fixed but unknown parameter cannot have one. The Bayesian definition treats probability as a measure of uncertainty, so a parameter, a unique event or a future observation can all have a probability distribution.

That difference has practical consequences. It is why a credible interval can say "the parameter is in here with 95% probability" and a confidence interval cannot, and why Bayesian models can predict new data directly. The examples on this page are made up for this app: a thyroid screening test, a treatment decision and a count of fish in a pond.

The page has two parts. **Part A** (concepts, no code needed) covers the ideas; mark each section as you go. **Part B** runs small R examples in your browser.

::: {.callout-note collapse="true"}
## Sources, reading and seminar
**Course notes:** Introduction and Module 1 "Bayesian concepts" (up to p. 52). Reading deadline: Seminar 1, Monday 7 September 2026.

**Optional textbook:** McElreath, *Statistical Rethinking* (2nd ed.), chapters 1–3 cover the same ground (grid approximation, sampling from the posterior).

**Examples on this page** (screening, treatment threshold, capture–recapture) are written for this app. The course notes use other examples; work through those in the notes.
:::

::: {.callout-tip collapse="true"}
## Before you start: three things from earlier courses
**Conditional probability.** $P(A \mid B) = P(A \cap B) / P(B)$. "Probability of A *given* B." In general $P(A \mid B) \neq P(B \mid A)$: the probability of a positive test given disease is not the probability of disease given a positive test.

**Law of total probability.** If $B$ takes the values $b_1, \dots, b_k$, then $P(A) = \sum_k P(A \mid b_k) P(b_k)$. This is the denominator of Bayes' rule.

**Likelihood.** For fixed data $y$, the likelihood is $p(y \mid \theta)$ viewed as a function of the parameter $\theta$. It is *not* a probability distribution over $\theta$ (it need not sum or integrate to 1). Maximum likelihood (STA110, STA140) picks the $\theta$ that makes the observed data most probable.
:::

# Part A · Concepts

## 1.1 Two meanings of probability {#t1-1 .topic flag="ilo"}

| | Frequentist (classical) | Bayesian |
|---|---|---|
| Probability is… | the long-run relative frequency of an event in infinitely many repetitions under identical conditions | a number between 0 and 1 that measures uncertainty (partial knowledge) about anything unknown |
| Can have a probability | repeatable events: the next coin toss, a random sample | also unique events and fixed unknown quantities: "this patient has the disease", "the true mean is above 5" |
| Parameters are… | fixed constants; randomness is only in the data | unknown, so described by a distribution |
| Typical output | estimator, standard error, p-value, confidence interval | posterior distribution, credible interval, posterior predictive distribution |

Two problems with the frequency definition: **"identical conditions" is vague** (the same coin, the same person tossing it, the same air pressure?) and, more seriously, **many events cannot be repeated even in principle**. Whether it rained in Gothenburg on 1 September 2026, or whether this particular patient responds to treatment, has no long-run frequency. Yet weather services publish a probability of rain every day, and clinicians talk about a patient's probability of disease. The Bayesian view makes that everyday usage formal.

::: {.callout-warning}
## Common confusion: "Bayesian means subjective, so it is less scientific"
Every analysis contains choices: which data to collect, which model to fit, which variables to adjust for, which results to report. The prior is one more choice, and it is written down explicitly where others can inspect and vary it. The course notes deliberately put the prior **last** in the list of error sources: data errors, software errors and model misspecification are usually more dangerous and harder to detect.
:::

:::: {.quiz}
Which statement can only be made within the Bayesian meaning of probability?

- The test will be positive in about 5% of healthy people
- **There is a 0.8 probability that this drug lowers blood pressure**
- If we resample, 95% of such intervals will cover the true mean
- The p-value is the probability of data this extreme under H0

::: {.exp}
"This drug lowers blood pressure" is a statement about a fixed unknown quantity (the true effect). Under the frequentist definition it is simply true or false and has no probability. The other three are statements about repeatable data-generating events.
:::
::::

## 1.2 Why use probability for beliefs? {#t1-2 .topic}

The notes give three justifications. None is a proof; each makes the Bayesian approach reasonable.

1. **Analogy.** We already describe physical randomness with probability. Using the same language for uncertainty due to incomplete knowledge is natural.
2. **Decision theory.** Decisions under uncertainty should minimise **expected loss**. If you want to act rationally on uncertain information, you need numbers that behave like probabilities.
3. **Coherence of bets.** If the prices you would pay for bets on events do not obey the rules of probability, a bookmaker can arrange bets so that you lose money whatever happens (a "Dutch book").

::: {.worked}
**Worked example: when to treat.** Suppose treating a healthy patient costs 1 unit (side effects, cost), and *not* treating a sick patient costs 9 units. Let $p$ be the probability that the patient is sick.

| Decision | Expected loss |
|---|---|
| Treat | $(1-p) \times 1$ |
| Do not treat | $p \times 9$ |

Treat when $(1-p) < 9p$, that is when $p > 1/10$. In general, treat when

$$p > \frac{\text{cost of a false positive}}{\text{cost of a false positive} + \text{cost of a false negative}}.$$

This threshold is the same for every patient. What differs is $p$, and $p$ is specific to *this* patient with *these* symptoms. A single long-run frequency cannot provide it; a probability for the individual case can.
:::

A decision that minimises expected loss can still turn out badly for a particular patient. Decision theory only promises the best average over repeated decisions made with the information available at the time.

:::: {.quiz}
A false negative costs 4 units and a false positive costs 1 unit. Above which probability of disease should you treat?

- 0.25
- **0.20**
- 0.80
- 0.50

::: {.exp}
Threshold = 1 / (1 + 4) = 0.20. Treat if P(sick) > 0.20.
:::
::::

## 1.3 Bayes' rule and the base-rate fallacy {#t1-3 .topic flag="ilo"}

$$p(B \mid A) = \frac{p(A \mid B)\, p(B)}{p(A)} = \frac{p(A \mid B)\, p(B)}{\sum_b p(A \mid B = b)\, p(B = b)}$$

Bayes' rule follows directly from the definition of conditional probability. Nobody disputes it. The debate between schools of statistics is about *which quantities may have probability distributions*, not about the rule itself.

::: {.worked}
**Worked example: screening for hypothyroidism.** In a screening population 2% have the condition. The test has sensitivity 0.95 (positive when sick) and specificity 0.90 (negative when healthy). A person tests positive. What is the probability that they have the condition?

Think in **natural frequencies**, 10,000 people:

| | Test positive | Test negative | Total |
|---|---|---|---|
| Condition | 190 | 10 | 200 |
| No condition | 980 | 8,820 | 9,800 |
| Total | **1,170** | 8,830 | 10,000 |

$$P(\text{condition} \mid +) = \frac{0.95 \times 0.02}{0.95 \times 0.02 + 0.10 \times 0.98} = \frac{190}{1170} \approx 0.16$$

Only 16%, despite a test that is right 95% and 90% of the time. There are more false positives (980) than true positives (190) because healthy people vastly outnumber sick people. Ignoring the prevalence (the prior) is the **base-rate fallacy**. Using Bayes' rule is uncontroversial here: the fallacy comes from *not* using it.
:::

Vocabulary that maps directly onto Bayesian terms:

| Diagnostic testing | Bayesian inference |
|---|---|
| Prevalence, pre-test probability | Prior $p(\theta)$ |
| Sensitivity and specificity | Likelihood $p(y \mid \theta)$ |
| Positive predictive value, post-test probability | Posterior $p(\theta \mid y)$ |

::: {.callout-warning}
## Common confusion: the p-value is not P(H0 | data)
A p-value is (roughly) $P(\text{data this extreme} \mid H_0)$. Reading it as $P(H_0 \mid \text{data})$ swaps the condition, the same mistake as reading sensitivity as the positive predictive value. Getting from one to the other requires Bayes' rule, and therefore a prior probability for $H_0$.
:::

:::: {.quiz}
The same test is used in a clinic where 20% of referred patients have the condition. What happens to the positive predictive value?

- It stays at 0.16, because the test is unchanged
- **It rises to about 0.70, as the prior is higher**
- It rises to 0.95, which is the test's sensitivity
- It falls, because more people are now tested

::: {.exp}
PPV = 0.95 × 0.20 / (0.95 × 0.20 + 0.10 × 0.80) = 0.19 / 0.27 ≈ 0.70. The likelihood (test accuracy) is unchanged; the prior (prevalence) changed, so the posterior changed.
:::
::::

## 1.4 A whole posterior from a discrete prior {#t1-4 .topic flag="ilo"}

Screening had one number as the answer. Usually the unknown can take many values and we want its **whole posterior distribution**. With a discrete parameter this still only needs multiplication and one sum.

::: {.worked}
**Worked example: how many fish are in the pond?** You catch, mark and release `r marked` fish. A week later you catch `r caught` fish and `r recaptured` of them carry a mark. Let $N$ be the number of fish in the pond.

**Likelihood.** If the pond has $N$ fish and fish mix freely, the number of marked fish in a catch of 25 follows a hypergeometric distribution:

$$p(y = 5 \mid N) = \frac{\binom{30}{5}\binom{N-30}{20}}{\binom{N}{25}}$$

It is zero for $N < 50$: you have already seen $30 + 20 = 50$ different fish. The maximum likelihood estimate is about $30 \times 25 / 5 = 150$ (the classical Lincoln–Petersen estimate).

**Prior.** A local angler says "a few hundred, probably not more than a thousand". A log-normal distribution with median 300 and log-scale SD 0.6 puts about 95% of its mass between roughly 90 and 1000. Restrict it to $N = 50, \dots, 3000$ and rescale so it sums to one.

**Bayes' rule on a grid.** For every candidate $N$: multiply prior × likelihood, then divide by the sum over all $N$.
:::

```{r}
#| label: fig-fish
#| fig-cap: "Prior (dashed), likelihood rescaled to sum to one (dotted) and posterior (solid) for the number of fish."
#| fig-height: 3.6
fish <- data.frame(
  N = rep(pond_size, 3),
  probability = c(prior, likelihood / sum(likelihood), posterior),
  curve = rep(c("prior", "likelihood (rescaled)", "posterior"), each = length(pond_size))
)
ggplot(fish, aes(N, probability, linetype = curve, colour = curve)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("grey40", "#1f5f99", "grey55")) +
  scale_linetype_manual(values = c("dotted", "solid", "dashed")) +
  coord_cartesian(xlim = c(0, 1200)) +
  labs(x = "number of fish in the pond (N)", y = "probability mass", colour = NULL, linetype = NULL)
```

The posterior sits between the prior (median 300) and the likelihood (peak near 150): posterior mode `r post_map`, median `r post_median`, mean `r post_mean`, and a 90% credible interval from `r ci_lower` to `r ci_upper`. The prior pulls the estimate up because five recaptures out of 25 is little data. The likelihood's long right tail (large ponds are hard to rule out with few recaptures) is tamed by the prior's belief that ponds of 2,000 fish are unusual.

Two conditions for this to work, from the notes' abbreviated workflow:

- the likelihood must be defined for every value the prior allows (it may be zero for some), and
- at least some values with non-zero prior must give the observed data non-zero probability. If the prior only allowed $N < 50$, no posterior would exist.

:::: {.quiz}
You rerun the fish example with a prior whose median is 1,000 instead of 300. What happens to the posterior?

- Nothing, because the data decide the posterior
- **It moves to larger N, since the data are weak**
- It moves to smaller N, towards the likelihood peak
- It becomes identical to the new, wider prior

::: {.exp}
With little data the prior has a large influence. A prior centred on 1,000 pulls the posterior upwards. More recaptures (more data) would make the likelihood narrower and reduce the prior's influence.
:::
::::

## 1.5 Summarising a posterior {#t1-5 .topic flag="ilo project"}

The posterior *is* the answer. Summaries are chosen for the reader and the decision at hand.

| Summary | Definition | Best under loss… |
|---|---|---|
| Posterior mean | $E[\theta \mid y]$ | squared error |
| Posterior median | 50th percentile | absolute error |
| Posterior mode (MAP) | $\arg\max_\theta p(\theta \mid y)$ | "all or nothing" (0–1 loss) |
| Posterior SD | SD of the posterior | describes uncertainty about $\theta$ |
| Equal-tailed 95% credible interval | 2.5th to 97.5th percentile | easy, works with transformations |
| 95% highest posterior density (HPD) interval | shortest region containing 95% | can be several pieces if the posterior has several modes |

**A 95% credible interval means what people usually want a confidence interval to mean:** given the model, the prior and the data, $\theta$ lies in the interval with probability 0.95.

**Transformations.** Because a monotone transformation keeps the order of values, **interval endpoints transform directly**: if $[-0.40, 0.10]$ is a 95% credible interval for a log relative risk, then $[e^{-0.40}, e^{0.10}] = [0.67, 1.11]$ is a 95% credible interval for the relative risk. **Means do not**: $E[e^{\theta}] \neq e^{E[\theta]}$. The median transforms directly because it is a percentile.

**With a sample of draws** (from Module 3 on, MCMC gives you draws from the posterior), every summary becomes an average or a percentile over draws. A derived quantity, such as a ratio of two parameters, is computed **draw by draw**, which keeps the correlation between parameters intact.

::: {.callout-warning}
## Common confusion: three different spreads
**Data SD** ($\sigma$): how much individual observations vary. **Posterior SD** of a mean: how uncertain you are about the mean; it shrinks as $n$ grows, like a standard error. **Posterior predictive SD**: how uncertain you are about one *new* observation; it never shrinks below $\sigma$. A larger $\sigma$ usually means larger posterior and predictive SDs too, but they answer different questions.
:::

:::: {.quiz}
A posterior for log(OR) has mean 0.40 and 95% equal-tailed interval [0.10, 0.70]. Which statement is correct?

- The posterior mean OR is exactly exp(0.40)
- **The 95% interval for OR is [exp(0.10), exp(0.70)]**
- The posterior SD of OR is exp of the SD of log(OR)
- Transforming an interval needs the delta method

::: {.exp}
Percentiles (and so equal-tailed interval endpoints and the median) survive monotone transformations. Means and SDs do not: E[exp(θ)] > exp(E[θ]) for a spread-out posterior. With draws you would simply compute exp() for every draw and summarise the result.
:::
::::

## 1.6 Posterior predictive distributions {#t1-6 .topic flag="project"}

Bayesian models are **generative**: given parameter values, you can simulate new data. Integrating over the posterior gives the **posterior predictive distribution**:

$$p(\tilde y \mid y) = \int p(\tilde y \mid \theta)\, p(\theta \mid y)\, d\theta$$

It is easiest to compute by Monte Carlo:

::: {.worked}
**Posterior predictive by Monte Carlo**

1. Repeat many times: (a) draw $\theta^{(s)}$ from the posterior; (b) draw a new $\tilde y^{(s)}$ from $p(y \mid \theta^{(s)})$.
2. Summarise the $\tilde y^{(s)}$: a mean, a probability such as $P(\tilde y \geq 10)$, or an expected loss.
:::

Fish example: next week you catch 40 fish. Each posterior draw of $N$ gives one simulated number of marked fish. Because $N$ is uncertain, the predictive distribution is wider than a hypergeometric distribution with $N$ fixed at its posterior mean. Plugging in a single best guess **underestimates uncertainty**.

To check whether you ran enough Monte Carlo draws, rerun the simulation and look at how much the answer changes: that variation is the **Monte Carlo error**. More draws shrink it at the rate $1/\sqrt{\text{number of draws}}$.

**Conditionally generative models.** A regression model predicts $y$ *given* $X$; it does not model the covariates themselves. That is fine when you choose $X$ (a dose, a treatment). **Not every method is generative**: a permutation test gives a p-value but cannot say what the next observation will look like.

:::: {.quiz}
Why is the posterior predictive distribution wider than the data model with θ fixed at its posterior mean?

- Because Monte Carlo sampling adds noise
- **It also includes the uncertainty about θ itself**
- Because the prior is added to the likelihood twice
- It is not wider; the two are identical

::: {.exp}
Predictive uncertainty = sampling variation of a new observation + uncertainty about the parameter. Fixing θ drops the second part.
:::
::::

## 1.7 The Bayesian workflow, abbreviated {#t1-7 .topic flag="project"}

The notes give a four-step version here and expand it in [Module 6](module6.qmd):

1. **Full probability model** for everything: observed data, unknown parameters, future or missing data. Requires knowing the research question and how the data were collected.
2. **Prior**: express the available information about the unknowns as a joint prior distribution.
3. **Condition on the data**: compute the posterior (and posterior predictive) of the quantities of interest.
4. **Evaluate**: does the model fit the data? Are the conclusions reasonable? How sensitive are they to the model and the prior?

Step 4 usually sends you back to step 1 or 2: the workflow is a loop, not a line. Modules 3 and 4 give the tools for step 4.

## 1.8 Confidence intervals are not credible intervals {#t1-8 .topic flag="ilo"}

Both are abbreviated "CI" and with large samples they often agree numerically. They answer different questions.

::: {.two-col}
::: {}
**95% confidence interval**

A property of the *procedure*: if the experiment were repeated infinitely often, 95% of the intervals produced would contain the true value. Says nothing about *this* interval.
:::
::: {}
**95% credible interval**

A property of *this* posterior: given the model, prior and data, the parameter lies in the interval with probability 0.95.
:::
:::

**Why a confidence interval cannot be read as a credible interval** (the notes' argument, with made-up numbers):

1. **Coverage is about the procedure, not the interval.** "Flip a coin: heads, report $(-\infty, \infty)$; tails, report the single point $\{0\}$" is a valid 50% confidence procedure for any non-zero parameter. It never uses the data. Its intervals are either useless or almost certainly wrong, yet the procedure has exact 50% coverage. So **the width of a confidence interval does not by itself measure precision**.
2. **Several valid procedures give different intervals for the same data.** With 4 successes out of 20:

```{r}
#| label: ci-table
x <- 4; n <- 20; p_hat <- x / n
wald  <- p_hat + c(-1, 1) * 1.96 * sqrt(p_hat * (1 - p_hat) / n)
wilson <- as.numeric(prop.test(x, n, correct = FALSE)$conf.int)
exact <- as.numeric(binom.test(x, n)$conf.int)
flat_cred <- qbeta(c(0.025, 0.975), 1 + x, 1 + n - x)
knitr::kable(data.frame(
  Interval = c("Wald 95% CI", "Wilson 95% CI", "Exact (Clopper–Pearson) 95% CI",
               "95% credible interval, Beta(1, 1) prior"),
  Lower = round(c(wald[1], wilson[1], exact[1], flat_cred[1]), 3),
  Upper = round(c(wald[2], wilson[2], exact[2], flat_cred[2]), 3)
))
```

   If every valid 95% confidence interval were a 95% probability statement about $\theta$, the region between the Wilson and exact limits would have to contain both 0% and a positive share of the probability. That is a contradiction. So confidence intervals are **not automatically** probability statements; at most one procedure could coincide with some posterior, and not *because* it is a confidence procedure.

3. **A credible interval has no such contradiction**: one prior + one model + one dataset gives exactly one posterior. Two people with different priors get different posteriors, just as two people with different data or models would.

**Do credible intervals have 95% coverage in repeated experiments?** Not in general. Coverage depends on how the prior relates to the true value: a prior that puts zero probability on the true value gives zero coverage. [Module 2, Section 2.8](module2.qmd#t2-8) shows this in a simulation.

::: {.callout-warning}
## Common confusion: "95% of the time, the true value lies in my confidence interval"
The true value is fixed and your interval is fixed: it either contains the value or not. The 95% refers to the long run of hypothetical repetitions. The sentence "there is a 95% probability that the effect lies between A and B" is a credible-interval statement and needs a prior.
:::

:::: {.quiz}
A colleague computed a 95% confidence interval [0.08, 0.42] and says "so the proportion is between 0.08 and 0.42 with 95% probability". What is the best reply?

- Correct, that is the definition of a confidence interval
- **That is a credible-interval statement; it needs a prior**
- Correct, as long as the sample size is above 30
- Only true for the exact interval, not for the Wald one

::: {.exp}
The 95% of a confidence interval refers to the procedure over repeated samples. A probability statement about the parameter itself requires a posterior, and so a prior. With a flat prior and a large sample the numbers are often close, which is why the confusion is common.
:::
::::

## 1.9 Hypothesis tests and Bayes factors {#t1-9 .topic}

You *can* compute $P(\theta > 0 \mid y)$ from a posterior. Two cautions from the notes:

- It depends on the prior probability that $\theta > 0$. If you then dichotomise at 0.05, two reasonable priors can give 0.049 and 0.051 and opposite decisions.
- For a continuous parameter $P(\theta = 0 \mid y) = 0$: a point null always has zero posterior probability.

The **Bayes factor** compares whole models instead:

$$\underbrace{\frac{p(m_1 \mid y)}{p(m_2 \mid y)}}_{\text{posterior odds}} = \underbrace{\frac{p(y \mid m_1)}{p(y \mid m_2)}}_{\text{Bayes factor}} \times \underbrace{\frac{p(m_1)}{p(m_2)}}_{\text{prior odds}}$$

BF > 1 favours $m_1$. The Bayes factor does not depend on the prior odds of the models, but it **does depend on the priors for the parameters inside each model**: a vague prior for an effect in $m_1$ spreads prior mass over huge effects and lowers $p(y \mid m_1)$. You must specify what the alternative hypothesis looks like. The course notes do not use hypothesis testing after Module 1. Later modules compare models by predictive performance instead ([Module 4](module4.qmd)).

## When does the Bayesian answer differ most? {#when-to-use}

- **Small samples and strong prior information**: the prior visibly moves the answer (the fish example).
- **Unique events and decisions**: a probability for *this* patient, *this* trial, or an expected loss.
- **Parameters near a boundary** (proportions near 0 or 1, variances near 0), where classical intervals can include impossible values.
- **With large samples and weak priors** the likelihood dominates and Bayesian and classical results usually agree numerically, though their interpretations still differ.

## Limitations {#limitations}

- The answer depends on the prior and the model. That dependence is visible and can be checked (sensitivity analysis), but it cannot be removed.
- Translating vague knowledge ("probably small") into a specific distribution is hard, and "I have no idea" cannot be expressed exactly (see flat priors in [Module 2](module2.qmd#t2-4)).
- A grid over all parameter values only works for one or two parameters. Beyond that, computation needs MCMC ([Module 3](module3.qmd)).

# Part B · Practice in the browser

::: {.callout-note}
## Interactive R
Cells marked *Run Code* run R in your browser. The first run on a page takes a few seconds while R loads.
:::

**Exercise 1 · Screening.** The function computes the positive predictive value.

```{webr-r}
#| label: ppv
ppv <- function(prevalence, sensitivity, specificity) {
  true_positive  <- sensitivity * prevalence
  false_positive <- (1 - specificity) * (1 - prevalence)
  true_positive / (true_positive + false_positive)
}
ppv(prevalence = 0.02, sensitivity = 0.95, specificity = 0.90)
```

1. What is the PPV in a general population where the prevalence is 0.2%?
2. Which prevalence gives a PPV of 0.5? Try values, or solve with `uniroot()`.
3. Keep prevalence at 2%. Which matters more for PPV here: raising sensitivity to 0.99 or raising specificity to 0.99?

::: {.callout-tip collapse="true"}
## Model answer
1. `ppv(0.002, 0.95, 0.90)` ≈ 0.019. Fewer than 2 in 100 positives have the condition.
2. `uniroot(function(p) ppv(p, 0.95, 0.90) - 0.5, c(0.001, 0.5))$root` ≈ 0.095, about 9.5%.
3. Specificity: `ppv(0.02, 0.99, 0.90)` ≈ 0.17 but `ppv(0.02, 0.95, 0.99)` ≈ 0.66. With a rare condition, false positives come from the large healthy group, so specificity dominates.
:::

**Exercise 2 · The fish posterior on a grid.**

```{webr-r}
#| label: fish-grid
marked <- 30; caught <- 25; recaptured <- 5
pond_size <- 50:3000                                   # every candidate N

prior <- dlnorm(pond_size, meanlog = log(300), sdlog = 0.6)
prior <- prior / sum(prior)                            # a proper prior on the grid
likelihood <- dhyper(recaptured, marked, pond_size - marked, caught)
posterior  <- prior * likelihood
posterior  <- posterior / sum(posterior)               # Bayes' rule: divide by the sum

post_cdf <- cumsum(posterior)
c(mode   = pond_size[which.max(posterior)],
  median = pond_size[which(post_cdf >= 0.5)[1]],
  mean   = sum(pond_size * posterior),
  lower90 = pond_size[which(post_cdf >= 0.05)[1]],
  upper90 = pond_size[which(post_cdf >= 0.95)[1]])

plot(pond_size, posterior, type = "l", xlim = c(0, 1200), lwd = 2,
     xlab = "N", ylab = "probability")
lines(pond_size, prior, lty = 2)
```

4. Replace the prior with a flat one (`prior <- rep(1, length(pond_size))`, then normalise). How do the mean and the 90% interval change? Why does the upper limit depend on the arbitrary cut-off at 3000?
5. Suppose `recaptured` had been 15. Rerun with the log-normal prior. Does the prior still matter much?
6. What is the posterior probability that the pond holds more than 400 fish?

::: {.callout-tip collapse="true"}
## Model answer
4. The mean rises (about 230) and the 90% interval widens to about 106–477. With a flat prior the long right tail of the likelihood is not tamed; with too few recaptures, very large ponds stay somewhat plausible. Change the grid to `50:10000` and the mean moves again: a "flat" prior on a bounded grid is still a choice.
5. With 15 recaptures the likelihood is concentrated near 50 fish, and prior and flat-prior posteriors almost agree. More data means less prior influence.
6. `sum(posterior[pond_size > 400])` ≈ 0.035.
:::

**Exercise 3 · Posterior predictive by Monte Carlo.** Run after Exercise 2 (it uses `posterior`).

```{webr-r}
#| label: fish-predict
n_draws <- 20000
N_draws <- sample(pond_size, n_draws, replace = TRUE, prob = posterior)  # step 1a
marked_next_week <- rhyper(n_draws, marked, N_draws - marked, 40)       # step 1b: catch 40

c(mean = mean(marked_next_week),
  prob_10_or_more = mean(marked_next_week >= 10))

# Naive version: plug in the posterior mean of N
N_plugin <- round(sum(pond_size * posterior))
naive <- rhyper(n_draws, marked, N_plugin - marked, 40)
c(sd_full = sd(marked_next_week), sd_plugin = sd(naive))
```

7. Why is `sd_full` larger than `sd_plugin`?
8. Run the cell three times. How much does `prob_10_or_more` change? What does that tell you about the Monte Carlo error with 20,000 draws?

::: {.callout-tip collapse="true"}
## Model answer
7. `sd_full` includes uncertainty about $N$ as well as the randomness of the catch; the plug-in version pretends $N$ is known.
8. It changes in the third decimal. That variation *is* the Monte Carlo error; it is small compared with the uncertainty the probability expresses, so 20,000 draws is enough here.
:::

## Why-questions {#why-questions}

::: {.callout-note collapse="true"}
## Why is Bayes' rule uncontroversial but Bayesian statistics debated?
Bayes' rule is a theorem of probability. The debate is about whether unknown *fixed* quantities (parameters, hypotheses, unique events) may have probability distributions, and so whether a prior may be used at all.
:::

::: {.callout-note collapse="true"}
## Why does the prior matter less as the sample grows?
The likelihood is a product over observations, so it becomes more and more concentrated while the prior stays the same. Eventually the likelihood dominates the product prior × likelihood. The exception: values with zero prior probability stay impossible whatever the data say.
:::

::: {.callout-note collapse="true"}
## Why should the posterior predictive not use a single best estimate?
A single estimate ignores the uncertainty about the parameter, so predictions are too narrow and rare events look rarer than they are. Tail events (extreme values, rare adverse outcomes) are where this matters most.
:::
